---
title: "Machine Learning & Prediction"
author: "Irene Yao"
date: "4/16/2018"
output:
  html_document:
    df_print: paged
---

Based on the [previous data analysis](https://github.com/ireneyaoyao/Springboard/blob/master/Capstone/Statistical%20Analysis.pdf), we will use the following variables for prediction of outcome type. 

* size
* intake_condition
* outcome_condition
* sex
* age_at_intake
* stage_at_outcome (age)
* days in shelter
* breed (is mix or not)

#### **Machine Learning and Prediction**
For this report's purpose, the prediction will be made around dogs, and the predicted value will be the outcome of each animal. 
I will use GBM for the prediction and the outcome will be a binary classification. The outcomes for the animals will be either "placed in a home" or "not placed in a home". "Adoption" and "Return to owner" will regarded as "placed in a home", and all others will be categorized into "not place in a home". 
To do so, I will add a column "placed" and the binary value for the column will be 1 or 0.

```{r}
dogs$placed <- ifelse((dogs$outcome_type == "ADOPTION" | dogs$outcome_type == "RETURN TO OWNER"), 1, 0)
```

Some of the columns have a class of "character" or "timediff". In order for the prediction model to work, update those columns to either "factors" or "numeric". 

```{r}
dogs$sex_clean <- as.factor(dogs$sex_clean)
dogs$stage_at_outcome <- as.factor(dogs$stage_at_outcome)
dogs$age_at_intake <- as.numeric(dogs$age_at_intake)
dogs$age_at_outcome <- as.numeric(dogs$age_at_outcome)
```


##### **Prediction Using GBM**
1. separate the dataframe into a training and a testing set. 80% of data will be in training set and the rest 20% will be in testing set. 

```{r}
n <- nrow(dogs)
n_train <- round(n * 0.8)
set.seed(123)
train_indices <- sample(1:n, n_train)
dog_train <- dogs[train_indices, ]
dog_test <- dogs[-train_indices, ]
```

2. create the GBM model

```{r warning=FALSE, message=FALSE}
library(gbm)
set.seed(1)
dog_model_gbm <- gbm(formula = placed ~ size + intake_condition + outcome_condition + sex_clean + age_at_intake + stage_at_outcome + is_mix,
                             distribution = "bernoulli",
                             data = dog_train,
                             n.trees = 10000)
```

3. predict the outcomes of the test set

```{r}
pred_gbm <- predict(object = dog_model_gbm, 
                newdata = dog_test,
                n.trees = 10000,
                type = "response")
```

4. evaluate the model using test set AUC

```{r warning=FALSE, message=FALSE}
library(Metrics)
auc <- auc(actual=dog_test$placed, predicted=pred_gbm)
print(paste0("Test set AUC: ", auc))
```



